---
title: "ironProjectAnnaWawer"
author: "Melanie Smith"
date: "13/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    fig.align = "center"
)
```

This RScript is the first in the collaboration with Anna and Burnd.  
Here I will inspect the STOP metadata and create a table of sample IDs suitable for Anna's Iron Infusion Project.  
The process of inclusion is as follows:  
  - Import all recorded data and data dictionary for the STOP project  
  - Subset the metadata to include only  
    - variables related to bloods, blood loss, vitamins, iron, pregnancy outcome  
    - women for whom we have frozen placenta samples  
    - women who match the required age and BMI range  

```{r include=FALSE}

###################################################
#                 Load Libraries                  #
###################################################

# Data manipulation
library(dplyr)
library(magrittr)
library(reshape2)
library(readr)
library(tibble)
library(readxl)

# Visualisation
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(knitr)
library(kableExtra)
library(plotly)

# file locations
library(here)

###################################################
#              Set ggplot theme                   #
###################################################

theme_set(theme_bw() +
            theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")))

###################################################
#                   Directories                   #
###################################################

projectDir <- "/home/smit1924/ironReceptor_placenta"

inputDir <- file.path(projectDir, "rawData")
outputDir <- file.path(projectDir, "cleanData")
dataFolder <- file.path(projectDir, "dataFolder")


###################################################
#             User Defined Functions              #
###################################################

# Negate the magrittr %in% function
`%notin%` <- Negate(`%in%`)
# REDCap import
source(file.path(projectDir, "RScripts", "redcap_api_import.R"))
```

## Iron analysis {.tabset}
### Import the metadata files

```{r import STOP}

# variables I want
STOP_variables <- c("participant_id", "f1_age",
                    "f1_lmp", "f4_ethnicity",
                    "f5_menarche", "f7_anaemia_yr", 
                    "f7_anaemia_med", "f9_wgt",
                    "f9_hgt", "f9_bmi", 
                    "f9_hb", "f9_ferritin",
                    "f9_b12", "f9_vitd",
                    "f21_est_blood_loss", "f22_gestw_del",
                    "f31_outcome_combination",
                    "f21_delivery_date_time")


# STOP (REDCap import)
STOP_raw_dat <- read.csv(file.path(
  dataFolder, "STOPStudy_DATA_2021-04-23_1151.csv"
  ),
                         header = TRUE)

STOP_raw_dict <- read.csv(file.path(
  dataFolder,"STOPStudy_DataDictionary_2021-04-23.csv"
  ),
                          header = TRUE)
# do the funky Shalem things
STOP_raw <- apply_data_dict(STOP_raw_dat,STOP_raw_dict)
table(STOP_raw$final_data) #use final data subset (N=1300) only

# add the gestational age for blood collection
GAsampling <- read_csv(file.path(dataFolder, "SCOPE+STOP_GAsamp.csv")) %>% 
  dplyr::filter(., Study == "STOP") %>% 
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(Study.ID, width = 4, pad = "0")
  ))

# subset for the variables we want
subsetSTOP <- dplyr::select(STOP_raw, final_data, all_of(STOP_variables)) %>%
  # drop samples not in the final data subset
  dplyr::filter(., final_data == "Yes") %>% 
  # sort by ID
  dplyr::arrange(., participant_id) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(participant_id, width = 4, pad = "0"))
    ) %>%
  dplyr::left_join(., GAsampling[, c("samplename", "GA_samp")],
                   by = "samplename") %>% 
  # make samplename the first column
  dplyr::select(., samplename, everything(), -final_data, -participant_id)

# Import the data with freezer sample info
freezerPlacenta <- read_csv(
  file = file.path(dataFolder, "STOP_placentaFreezerSamples.csv")) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(Patient_ID, width = 4, pad = "0")))
# drop samples that have been removed from the freezer
freezerPlacenta <-  dplyr::filter(freezerPlacenta, sampleRemoved == 0)

# remove women with missing data
secondSubsetSTOP <- subsetSTOP[complete.cases(
  subsetSTOP[, c("f9_ferritin", "f9_hb")]), ] # %>% 
  # remove women outside of the maternal age, BMI or gestational age range
  # I've given us 10% either side to broaden the range
  # dplyr::filter(., between(f1_age, 19*.9, 40*1.1)) %>% 
  # dplyr::filter(., between(f9_bmi, 20.6*.9, 51.1*1.1)) %>% 
  # dplyr::filter(., between(f22_gestw_del, 37*.9, 41.6*1.1))

# remove women without frozen placenta tissue
thirdSubsetSTOP <- subset(secondSubsetSTOP,
                          samplename %in% unique(freezerPlacenta$samplename))

# Add column for samples we alreay have sequencing for
# import PE sample data
PE_placenta_meta <- read_csv(
  file.path(dataFolder, "PlacentaGroup_Sample_information.csv")) 
# keep the STOP IDs
STOP_PE_placenta_meta <- dplyr::filter(PE_placenta_meta, cohort == "STOP")

# import GDM sample data
GDM_placenta_meta <- read_csv(
  file.path(dataFolder, "GDM_metadata_cleaned.csv")) 
# add a samplename column that matches the rest of my data
GDM_placenta_meta %<>% dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(ID, width = 4, pad = "0")))
# keep the STOP IDs
STOP_GDM_placenta_meta <- dplyr::filter(GDM_placenta_meta, Cohort == "STOP")

# add column to the metadata 0=not sequenced before; 1=sequenced before
fourthSubsetSTOP <- thirdSubsetSTOP %>%
  dplyr::mutate(., sequencedBefore = ifelse(
    thirdSubsetSTOP$samplename %in% 
      c(STOP_GDM_placenta_meta$samplename, STOP_PE_placenta_meta$samplename),
    1, 0))

# import the UR conversion table
convertSTOP_UR <- read_csv(file.path(dataFolder, "convertSTOP_UR.csv")) %>% 
  # add column with samplename in proper format
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(STOP_ID, width = 4, pad = "0")
  ))

# new DF including the patient UR
fourthSubsetSTOP_UR <- dplyr::left_join(
  fourthSubsetSTOP,
  convertSTOP_UR[, c("maternalUR", "samplename")],
  by = "samplename") %>% 
  # rename and reorder columns
  dplyr::select(., samplename, maternalUR, sequencedBefore,
                maternalAge = f1_age, 
                gestationalAgeDelivery = f22_gestw_del,
                gestationalAgeBloodTest = GA_samp, 
                f9_ferritin, f9_hb, f9_b12, f9_vitd,
                everything()) %>% 
  dplyr::arrange(., sequencedBefore)

# write the table to file for Anna
library("writexl")
# write_xlsx(fourthSubsetSTOP_UR,
#            here("cleanData/fourthSubsetSTOP_UR.xlsx"))

```

### data dictionary table

```{r data dictionary table}

STOP_subsetDataDic <- subset(
  STOP_raw_dict, Variable...Field.Name %in% STOP_variables)

STOP_subsetDataDic %>%
  # head(50) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling() %>%
  kableExtra::scroll_box(height = "600px")

```

### metadata table final subset

<br/>
- The STOP cohort has `r dim(secondSubsetSTOP)[1]` women inside the maternal age, maternal BMI and gestational age at delivery window. Of these, `r dim(thirdSubsetSTOP)[1]` have frozen tissue available. Note that `r dim(dplyr::filter(fourthSubsetSTOP, sequencedBefore == 1))[1]` tissue samples have been sequenced previously, leaving `r dim(dplyr::filter(fourthSubsetSTOP, sequencedBefore == 0))[1]` unsequenced samples.
<br/>

```{r metadata table fourth subset}

fourthSubsetSTOP %>%
  # head(50) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling() %>%
  kableExtra::scroll_box(height = "600px")

```